---
title: "CytofIN Tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{my-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# CytofIN

CytofIN (CyTOF integration) is an R package for homogenizing and integrating heterogenous CyTOF data from diverse data sources.

Before CyTOF data integration, all CyTOF files need to be homogenized to have consistent channels. CytofIN requires that all input CyTOF files be homogenized based on a user-provided standardized panel with user defined search pattern. To normalize the CyTOF data, CytofIN uses a novel generalized anchor strategy that defines the based line of the signal between batch to correct for batch effects. One anchor needs to be identified by the user from each plate (batch). A reference anchor is generated based on the mean expression of all identified anchors from each plate (batch). Next, a user-specified transformation function is applied to fit each plate-specific anchor to the reference data distribution and the same transformation is then applied to correct the sample data signal on each plate.  

CytofIN provided three functions for CyTOF data integration:

1. homogenize-this function performs batch homogenization of CyTOF data based on a user-defined panel and search pattern. 
2. anprep-this function generates reference statistics from anchors identified from each plate (batch).
3. annorm-this function performs signal normalization using transformation function based on the anchor from anprep.
4. annorm_nrs-this function performs signal normalization using stabilized channels as internal anchors. 

```{r}
library(knitr)
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  lines <- options$output.lines
  if (is.null(lines)) {
    return(hook_output(x, options))  # pass to default hook
  }
  x <- unlist(strsplit(x, "\n"))
  more <- "..."
  if (length(lines)==1) {        # first n lines
    if (length(x) > lines) {
      # truncate the output, but add ....
      x <- c(head(x, lines), more)
    }
  } else {
    x <- c(more, x[lines], more)
  }
  # paste these lines together
  x <- paste(c(x, ""), collapse = "\n")
  hook_output(x, options)
})
``` 


```{r setup}
library(devtools)
install_github('bennyyclo/Cytofin')
```

**CyTOF data homogenization**
```{r, echo=FALSE, out.width="100%"}
img1_path <- "../images/Slide1.png"
knitr::include_graphics(img1_path)
```

Description:
the homognize function takes a user input antigen panel table, which includes standardized antigen name and associated antigen search pattern. Given two CyTOF files with distinct antigen naming, the program performs a regular expression search to match the synonymous term in the panel and correct the antigen name with standardized names in the panel.

Function Definition: 

```homogenize(metadata_filename, panel_filename, input_file_dir, output_file_dir)```

Input: 

```metadata_file```: metadata table of raw CyTOF files (.fcs)(must be in the current directory).

```panel_filename```: standardized antigen panel table file (.xlsx/.csv)(must be in the current directory).

```input_file_dir```: folder directory containing input raw CyTOF files.

```output_file_dir```: folder directory containing output homogenized files.

Output: homogenized CyTOF file with user-defined channels presented in the standardized antigen table.

$~$

**CyTOF data normalization using external anchors**

```{r, echo=FALSE, out.width="100%"}
img2_path <- "../images/Slide2.PNG"
knitr::include_graphics(img2_path)
```

The external anchor normalization steps include: 1. preparation of external anchors and 2. application of transformation function.

$~$

1. Anchors preparation:

Description: 
the anprep function concatenates the identified anchor file, one file per plate/batch, and subsequently generates summary statistics including mean and variance which will be used for batch correction. 

Function definition: 

```anprep(metadata_filename, panel_filename, input_file_dir)```

Input: 

```metadata_filename```: metadata table of anchor CyTOF files (.fcs)(must be in the current directory).

```panel_filename```: standardized antigen panel table file (.xlsx/.csv)(must be in the current directory).

```input_file_dir```: folder directory containing output data.


Output: an RData object containing reference statistics of the universal reference and concatenated anchor FCS files.

The RData object stored the following variables regarding the universal reference. The variables are exported from the RData object for subsequent batch normalization:

```mean_uni```: a 1-dimensional array of mean expression for all input markers 

```mean_var```: a 1-dimensional array of mean variance values of all marker expressions. 

```mean_uni_mean```: the mean value of mean_uni array.

```mean_uni_var```: the mean value of mean_var array.

$~$

2. Data transformation:

Description: the annorm function applied different transformation functions (modes) to normalize each anchor to the referenece statistcs generated by the anprep function.

Function definition:

```annorm (control_metadata_filename, control_data_filename, sample_metadata_filename, panel_filename, input_file_dir, val_file_dir="none" ,output_file_dir, mode)```

Input: 

```control_metadata_file```: metadata table of anchor CyTOF files (.fcs)(must be in the current directory).

```control_data_filename```: RData object containing anchor referene statistics (must be in the current directory).

```sample_metadata_filename```: metadata table of homogenized CyTOF files (.fcs)(must be in the current directory).

```panel_filename```: standardized antigen panel table file (.xlsx/.csv)(must be in the current directory).

```input_file_dir```: folder directory containing input homogenized CyTOF data file.

```val_file_dir```: folder directory containing validation homogenized CyTOF data file (optional).

```output_file_dir```: folder directory containing output normalized CyTOF data file.

```mode```: transformation function used for normaliztion (_meanshift_, _meanshift_bulk_, _variance_, _z_score_, _beadlike_).

 
Output: normalized CyTOF files.

$~$

**CyTOF data normalization using internal anchors**

```{r, echo=FALSE, out.width="100%"}
img3_path <- "../images/Slide3.PNG"
knitr::include_graphics(img3_path)
```

Description:
In the event that the external references are not available, internal anchors can be used. Here, we identified the most stable channels as internal anchors using a PCA-based non-redundnacy score (NRS). A minimal of three channels should be selected to establish an internal refernece from which signal can be calibrated between CyTOF files.

Function definition:

```annorm_nrs(sample_metadata_filename, panel_filename, input_file_dir, val_file_dir="none", output_file_dir, nchannels)```

Input: 

```sample_meta_filename```: metadata table of homogenized CyTOF files (.fcs)(must be in the current directory).

```panel_filename```: standardized antigen panel table file (.xlsx/.csv)(must be in the current directory).

```val_file_dir```: folder directory containing validation homogenized CyTOF data file (optional).

```output_file_dir```: folder directory containing output normalized CyTOF data file.

```nchannels```: number of stabilized channels used for normalization.


Output: normalized CyTOF files.

$~$

**Computational pipeline for CyTOF data integration**

Below is a demo Rscript using CytofIN package for CyTOF data integration.

First, let's homogenize the panel of all FCS files using the homogenize function:

```{r, out.width="100%", , output.lines=200}
#import cytofin R package
library(cytofin)

#homogenization antigen panel, use the demo data supplied with the package
metadata_filename <- paste0(path.package("cytofin"),"/extdata/test_metadata_raw.csv")
panel_filename <- paste0(path.package("cytofin"),"/extdata/test_panel.csv")
input_file_dir_homogenize <- paste0(path.package("cytofin"),"/extdata/test_raw_fcs_files/")
output_file_dir_homogenize <- "out_test/"
homogenize(metadata_filename, panel_filename, input_file_dir_homogenize, output_file_dir_homogenize)
```

This step homogenized the marker names of your target samples to the names indicated in your test_panel.csv file. 

Next, we will generate an RData object which contains statistics of the generalized anchors. 

```{r, out.width="100%"}
#prep external anchor 
anchor_metadata_filename <- paste0(path.package("cytofin"),"/extdata/test_anchor_metadata_raw.csv")
input_file_dir_anprep <- output_file_dir_homogenize #use the homogenized files
anprep(anchor_metadata_filename, panel_filename, input_file_dir_anprep)
```


Time to perform the batch normalization using CytofIN normalization function using healthy control samples as generalized anchors:

```{r, out.width="100%"}
#data normalization using external anchors and meanshift transofmration function
val_file_dir <- paste0(path.package("cytofin"),"/extdata/test_batch_fcs_files/")
anchor_data_filename <- "./Prep_control.RData"
output_file_dir_annorm <- "norm_test/"
mode <- "meanshift"
annorm(anchor_metadata_filename, anchor_data_filename, metadata_filename, panel_filename, 
input_file_dir_anprep, val_file_dir, output_file_dir_annorm, mode)
```

You can see samples get better after normalization when normalized signal (dark green) moved closer to the validation signal (purple). 

Now let us try to normalize with stable channels as generalized anchors:

```{r, out.width="100%"}
#data normalization using 4 internal channels and meanshift_bulk transformation function
nchannels <- 4
output_file_dir_annorm_nrs <- "norm_test2/"
annorm_nrs(metadata_filename, panel_filename, input_file_dir_anprep, val_file_dir, 
output_file_dir_annorm_nrs, nchannels)

```
